No Ramp Needed: Spandrels, Statistics, and a Slippery Slope

Version 1

Abstract

Summary

Previously, Tuller et al. found that the first 30 to 50 codons of the genes of yeast and other eukaryotes are slightly enriched for rare codons, so are presumably translated somewhat slowly. They argued, based on informatics, that this initial slow translation “ramp” was adaptive; and that it increased the efficiency of translation by provided a queuing mechanism to prevent ribosome collisions. Today, the translational speeds of the various codons are known, and indeed rare codons are translated slowly. We re-examined the slow translation ramp. We confirm the finding of Tuller et al. that the 5’ region of genes is enriched for rare codons. However, we also find that the 5’ ends of yeast genes are poorly conserved in evolution, suggesting that they are unstable and turn over relatively rapidly. When a new 5’ end forms de novo, it is likely to include codons that would otherwise be rare. Because evolution has had a relatively short time to select against these codons, 5’ ends are typically slightly enriched for rare, slow codons. Opposite to the expectation of Tuller et al., we show by direct experiment that genes with slowly translated codons at the 5’ end are expressed relatively poorly, and substituting faster codons improves expression. Further informatic studies suggest that for natural genes, slow 5’ ends are correlated with poor gene expression, opposite to the expectation of Tuller et al. Thus we conclude that slow 5’ translation is a “spandrel”; it is a non-adaptive consequence of something else, in this case the turnover of 5’ ends in evolution.

Introduction

Tuller et al. 1 were interested in the idea that a slow translational ramp at the beginning of a gene might queue ribosomes in an orderly way, thereby preventing ribosome traffic jams and collisions. However, at that time, translation speeds for the 61 sense codons were not known from direct measurement. Therefore, as a proxy for codon translation speed, Tuller et al. devised a proxy speed measurement based on the tRNA-adaptation index (tAI), a measure of the abundance of each tRNA. The assumption is that codons recognized by more abundant tRNAs would be translated faster. Using this proxy, Tuller et al. found that in yeast and other eukaryotes, the first 30 to 100 codons are enriched for codons for which tRNAs are rare (generally, rare codons), and are presumably translated slowly. The size of the effect is small (about a 3% difference, Tuller et al. Figure 2C), but is statistically significant.

At the time of the work of Tuller et al., ribosome profiling had recently been developed 2, and early ribosome profiling results showed a very high density of ribosomes near the 5’ end of the mRNA, consistent with very slow translation in this region, which could have been partly explained by the results of Tuller et al. However, later work showed that this 5’ high density of ribosomes was an artefact of the way cycloheximide was used to arrest translation in the original protocol3. With newer protocols for ribosome profiling, the region of 5’ high ribosome density largely disappears (but it does not entirely disappear3 (see Discussion)).

Since then, many workers have used ribosome profiling to directly measure the speed of translation of individual codons (cited below). With such data in hand, we re-visited the issues addressed by Tuller et al. 1. On the one hand, our analyses confirm that the 5’ regions of genes are typically slightly enriched for rare codons, and these encodings likely slow translation. On the other hand, various aspects of the data led us to an alternative hypothesis, namely that the 5’ ends were turning over relatively rapidly in evolution; that these 5’ ends were therefore relatively young; and that selection had not yet succeeded in removing all the rare, slow codons initially present in the de novo 5’ ends. We did a direct experimental test of the effects of slow or fast initial translation. Exactly opposite to Tuller et al., we found that encoding slow initial translation resulted in lower protein production than fast initial translation. This continued to be true even when we placed ribosome collision sites inside the reporter gene. Thus a slow initial translation ramp, though present, does not improve gene expression.

It is natural to assume that the enrichment of slow codons near 5’ ends is a product of selection. However, as elegantly argued by Gould and Lewontin4 in their classic paper “The Spandrels of San Marco and the Panglossian Paradigm: A Critique of the Adaptionist Programme”, not all biological phenomena are adaptive, or even a direct product of selection. They argued from the example of a “spandrel”: in architecture, a triangular space created when an arch supports a lintel. There is no architectural role for spandrels as such; they are the indirect and inevitable result of the juxtaposition of two other functional architectural elements. We argue that the slow initial translation of eukaryotic genes may likewise be a spandrel, a non-adaptive consequence of something else, the instability of 5’ ends in evolution.

Results

Calculations of encoded translation speed imply slow initial translation

Tuller et al. 1 used codon-specific tRNA abundance as a proxy to estimate speed of translation of codons. Since then, analysis of ribosome profiling data has yielded direct measurements of the translation speed of individual codons3,5–11. Accordingly, we have repeated some of the work of Tuller et al., but using the Ribosome Residence Time (RRT, Methods and Materials, Table S1) 6 of each of the 61 sense codons as a measure of translation speed. We refer to “encoded translation speed” to specify that we are focusing purely on the effects of different codons on translation speed, and not on other factors that might differentially affect translation speed at different regions of the mRNA.

Using the RRT, encoded translation speeds were calculated in sliding windows across all coding ORFs from S. cerevisiae. (The start codon was omitted because it is constant across genes and is an unusually “fast” codon.) Consistent with Tuller et al., we find that the first 30-100 codons had significantly lower calculated translation speeds than the rest of the gene (Figure 1). We focused our analyses on the first 40 codons for comparability to Tuller et al.; we call this the “Slow Initial Translation” region, or SIT. Although the tendency towards slow translation near the beginning of the gene is very highly statistically significant, the size of the effect is small. When we compare the first 40 codons of a gene to the rest of the same gene, we find that the average difference in encoded translation speed is about 1.2% (similar to 1), with a p-value < 0.001. For comparison, codons can vary in translation speed by about two-fold (Table S1), or perhaps as much as 6-fold (Weinberg et al., their Table S2, also shown in our Table S1).

Calculation of translation speed confirms slow initial translation (SIT). Translation speeds were calculated using Ribosome Residence Time (RRT) (Gardin et al. 2014) (Table S1 for RRT values) as a measure of codon-specific translation speed over 5694 S. cerevisiae ORFs. The horizontal line indicates average inverse RRT across all ORFs. The average speed in the first 40 amino acids is about 1.1% slower than in the rest of the gene (p < 0.001)

We observed a qualitatively similar, but smaller, encoded slow translation at the 3’ ends of (Figure S1).

(Note: Supplementary Figures and Tables are at the end of the manuscript, after References.)

Although on average genes are translated slowly near their 5’ ends, there is variability. Figure S2 shows the distribution of initial encoded translation speed for all yeast ORFs. About 57% of genes have a Slow Initial Translation (SIT) region, while the remainder have Fast Initial Translation (FIT).

Rare (slow) codons are enriched within the first 40 codons

Rare codons tend to be slow codons, and vice versa6 (Figure S3). But the correlation is not perfect—other things being equal, A/T-rich codons tend to be faster than G/C-rich codons, and codons with a wobble base at the third position tend to be faster than codons with the cognate base6. To better understand why gene beginnings are more slowly translated, we examined the relative usage of each of the 61 sense codons in the first 40 codons (Figure 2).

Codon usage in the Slow Initial Translation (SIT) region. A. Relative usage of each codon in the SIT versus the rest of the gene. The Y-axis shows the usage of each codon in the first 40 amino acids (omitting ATG) divided by its usage in the rest of gene. Each of the 61 sense codons is displayed along the X-axis, grouped by amino acid. Within each group, codons are ordered, left to right, from least frequent to most frequent over the whole transcriptome (i.e., the leftmost codon in each group is the rarest). Red arrows show the seven slowest codons by RRT, purple arrows show the seven rarest codons by total usage over the transcriptome, and blue shows Start and possible Start codons (ATG, ATA, ATC). Ratios above 1 indicate enrichment in the first 40 amino acids. Typically, the rarest codons for each amino acid are enriched. B. Absolute usage of each leucine codon in the SIT. The absolute usage frequency of each leucine codon is shown globally, and for the first 40 amino acids. Rare codons are still rare in the SIT, just not as rare as elsewhere. The same pattern holds for the other amino acids.

In the first 40 codons, we saw notable depletion of AUG (by nearly 50%), ATT, and ATC (Figure 2A). These are all fast codons, and their depletion would lead to slower translation. But AUG is the usual Start codon, and ATT and ATC can also be used as Start codons. Possibly these codons are depleted to reduce the possibility of translation initiating at the wrong place. We considered whether the SIT might be due to depletion of Start codons. We recalculated translation speeds after assigning AUG a neutral speed. On the one hand, this did reduce the difference in translation speed between the SIT and the rest of the gene by about 13% of the difference, but even so, the slow initial translation was highly significant. Thus, bias against Start codons contributes only modestly to slow translation.

We also saw enrichments of all six arginine codons (very likely because of their use in N-terminal signal sequences, data not shown), especially the slow, rare codons CGA, CGC, and CGG. Slow, rare codons for proline (CCC, CCG), leucine (CTC), glycine (GGG) and cysteine (TGC) were also enriched (Figure 2A).

Rare codons are rare in the first 40 codons, just not as rare as elsewhere

Although the proportion of rare, slow codons in the SITs is relatively higher than the proportion of rare codons in the body of genes, in absolute terms rare codons are still rare compared to more common synonymous codons (e.g., Figure 2B, Leu codons). That is, rare, slow codons are still strongly disfavored in the SIT, though they are less strongly disfavored than elsewhere. This was true for all rare codons.

Why are there relatively more rare, slow codons at 5’ ends?

Our analysis is consistent with that of Tuller et al. to the extent that we find a slight relative enrichment of rare, slow codons near the 5’ ends of coding regions. Tuller et al. interpret the slow translation ramp as an adaptation—they believe there is selection for slow codons near the 5’ end to enhance efficiency of translation. But there are other possibilities.

The Young Spandrel Hypothesis

We have noticed (see below) that the N-termini of yeast genes are often poorly conserved, and otherwise highly homologous genes often vary at the N-termini between different closely-related species. This suggests a different idea: N-termini are unstable and variable in evolution. They form de novo from new DNA sequence, and so all codons may initially occur at similar frequencies. Since these N-termini are, on average, younger than the remainder of the gene, selection has worked on them for a shorter time. Therefore selection against rare, slow codons may be less complete, and N-termini, due to their relative youth, may still retain some extra rare codons.

In this idea, in contrast to Tuller et al., the slight excess of rare codons near N-termini is not adaptive; it is not at all a product of selection. Instead, in the words of Gould and Lewontin 4, it is a spandrel. It is a non-adaptive by-product of something else, in this case the evolutionary instability of N-termini. This idea is explored below.

Poor 5’ conservation is a feature of many yeast genes

To illustrate the idea, we picked example genes. We used protein-protein BLAST at NCBI to blast several query genes against species of the subphylum Saccharomycotina (but having subtracted out all of Saccharomyces). Figure 3 shows that for these chosen example genes, the middle portions of the proteins are highly conserved, but the N-termini are not. We suggest that these non-conserved N-terminal regions are young in evolution, and therefore likely contain an excess of rare codons.

The N-termini of proteins can vary in evolution. BLAST of four example S. cerevisiae proteins against proteins in the subphylum “Saccharomycotina” (taxid: 147537) (excluding Saccharomyces, taxid 4930) was performed. Top hits are shown. Red regions indicate homology with an alignment score > 200, while white indicates no detected homology (BLAST default parameters). Even though all hits have high to moderate homology towards the center of the protein, many have little or no homology at the N-terminus.

In the next section, we ask whether the findings of Figure 3 can be generalized to proteins of S. cerevisiae.

Method of scoring N-terminal conservation, and rationale for using Saccharomycotina

We continued our investigations into N-termini, protein conservation, and translation speed using a more systematic approach. We ran local protein-protein BLAST for all S. cerevisiae genes against a database consisting of the Saccharomycotina subphylum, omitting Saccharomyces cerevisiae. Saccharomycotina was chosen because almost every gene from S. cerevisiae has a recognizable, conserved homolog in almost every species in Saccharomycotina, and yet the evolutionary distances are long enough that there is considerable sequence variability. We excluded species of Saccharomyces as they are too closely related to S. cerevisiae, and they are also very numerous in the database, and would overwhelm results from the other members of Saccharomycotina. We believe, however, that this particular choice of subphylum does not greatly affect the final result.

Conservation at the N-terminus was calculated as the weighted proportion of yeast species with sequence matches (a match by default BLAST parameters) beginning in the first 40 amino acids. The lowest conservation score is 0 (no hits in the first 40 amino acids), whereas the highest conservation score is 40 indicating that every species had a match (default BLAST parameters) starting at the first amino acid (Table S2, S3). The length of genes is negatively correlated with the N-termini conservation score (rho= −0.47; p<0.001)—that is, short genes tended to have more conserved N-termini.

N-termini are variable and poorly conserved

We measured protein conservation across more than 3,000 S. cerevisiae proteins with orthologues among 822 closely related yeasts from Saccharomycotina. For each protein, we developed a conservation score for the first 40 amino acids to represent the N-termini, an equivalent conservation score for the middle 40 amino acids, and an equivalent conservation score for the C-terminal 40 amino acids (Methods and Materials; Tables S2, S3).

Strikingly, the N-termini of S. cerevisiae orthologs had conservation scores that were much lower, and very differently distributed than the middle of the same orthologs (Figure 4). The first 40 amino acids had a flat distribution of protein conservation scores, indicating high levels of variability amongst these orthologs. That is, many of the orthologs had no detectable homology with the first 40 amino acids of the cerevisiae protein.

Conservation of S. cerevisiae proteins over the N-terminal, Middle, and C-terminal 40 amino acids. S. cerevisiae proteins were blasted against proteins of Saccharomycotina (excluding cerevisiae). “Conservation Scores” (Methods and Materials) were calculated for the N-terminal, Middle, and C-terminal 40 amino acids of the S. cerevisiae proteins. Scores range from 0 (no conservation) to 40 (perfect conservation). The frequency of each conservation score (3964 S. cerevisiae proteins) was plotted.

In contrast, the middle 40 amino acids were highly conserved, with conservations scores peaking at 40. It Is evident that for the middle 40 amino acids, a large fraction of orthologs had a region of high homology to the S. cerevisiae protein, whereas this was not true for the N-termini. Finally, the last 40 amino acids had conservation scores similar to those of the first 40 amino acids, though a bit higher (more conserved) (see Figure S4 for a comparison). These results suggest both ends of the gene “breathe”, gaining and losing new sequences during evolution, whilst the middles of the same genes stay constant. Thus the ends of the genes are younger than the middles. At their first formation, they would likely have contained some rare codons, which selection may not yet have had time to remove.

5’ translation speeds positively correlate with 5’ conservation scores

If the “Young Spandrel” hypothesis is true, and slow 5’ translation is caused by evolutionary instability of 5’ ends, then there should be a correlation between encoded slow translation, and poor N-terminal conservation. Our model predicts the newest N-termini to have the lowest conservation and the rarest codons, and, vice versa, the termini with the rarest codons should have the lowest conservation scores. To test this, we ranked all genes by the conservation scores of their first 40 amino acids. We then divided this ranked list into thirds. For each of the thirds (i.e., the bottom, middle, and top conservation scores) we plotted the average relative initial translation speeds.

As shown in Fig. 5A, the genes with the most poorly conserved N-termini also had the slowest initial translation, while the genes with the most conserved N-termini had the fasted initial translation, supporting the Spandrel hypothesis, and opposite to the Ramp hypothesis.

Slow Initial Translation is correlated with poor N-terminal conservation. A. Proteins were grouped by their N-terminal conservation scores (top, middle, and bottom thirds), and then the relative initial translation rate was plotted for each group. More conserved N-termini have faster initial translation. B. Proteins were grouped by their initial translation rate (top, middle, and bottom thirds), and then the N-terminal conservation scores were plotted for each group. Genes with faster initial translation have more conserved N-termini. Relative Initial Translation Speed is the log2 of (average RRT of the first 40 amino acids divided by the average RRT of the rest of the same gene) (Methods and Materials).

We also looked at the correlation in the other direction (Fig. 5B). We ranked genes by their relative initial translation speed, and divided the ranked list into thirds, then plotted N-terminal conservation scores. Again the effects are correlated: genes with the slowest initial translation have the lowest N-terminal conservation scores. Thus, overall, there is a strong correlation between N-terminal instability (i.e., newness in evolution, low conservation scores) and slow initial translation (i.e., the presence of slow/rare codons).

The Ramp hypothesis is inconsistent with observations of ribosome density and gene expression

In the Tuller “Ramp” hypothesis, in which the purpose of the slow translational ramp is to queue ribosomes and prevent collisions, genes with the highest ribosome occupancy would be in the most danger of ribosome collisions, and would therefore presumably have pronounced SITs. SITs might not be necessary on genes with low ribosome density, since there would not be much danger of collision in any case. To test this, we used the Arava et al. 2003 dataset which measured the density of ribosomes on all S. cerevisiae mRNAs12 (Figure 6). We ranked genes by ribosome density, then grouped them in thirds. Opposite to the expectation from the Tuller et al. “Ramp” theory, the genes with the highest ribosome densities had the fastest initial translation, whereas the genes with the lowest ribosome densities had the slowest initial translation (Figure 6C). While these findings are opposite to the expectation of the “Ramp” theory, they are consistent with the spandrel theory, because genes with high ribosome density would be subject to more intense selection against slow codons, thus leading to faster 5’ ends. Furthermore, analysis of Conservation Scores on the same genes showed that the genes with the lowest ribosome densities also had the lowest Conservation Scores (Figure 6D), as predicted by the Spandrel hypothesis.

Genes with high levels of expression, and high ribosome densities, generally have rapidly-translated N-termini, and high N-terminal conservation scores. A and B. Genes were grouped by expression level (bottom, middle, and top)(except that genes with fewer than 10 read-counts were omitted to reduce noise) (Lipson et al. 2009). In A, the initial translation rate is shown; in B, the conservation scores are shown. The correlation between speed and transcript abundance fails for the bottom third of genes; possibly these are genes expressed at high levels under other conditions (e.g., meiosis and sporulation). C and D. Genes were grouped by ribosome density (Arava et al. 2003) as a measure of intensity of translation. In C, the initial translation rate is shown; in D, the conservation scores are shown. High ribosome density correlates with high initial translation speed and high conservation score.

Similarly, in the “Ramp” hypothesis, ribosome collisions on highly-expressed genes would presumably have more serious consequences for the cell than collisions on poorly-expressed genes, and so highly expressed genes ought to have the most pronounced SITs. To test this, we used the Lipson et al. 2009 transcriptome dataset which measured the number of mRNA transcripts for S. cerevisiae genes13. Again, we ranked genes by expression, then grouped them by thirds. Exactly contrary to the “Ramp” hypothesis, we found that genes with the highest expression had the fastest initial translation (Figure 6A). In this analysis, the middle and bottom genes are not significantly different from each other; possibly some of the poorly expressed genes are inducible genes that would be highly expressed under some other condition (e.g., the sporulation genes, the GAL genes). In addition, the most highly expressed genes had the highest conservation scores, consistent with the Spandrel hypothesis (Figure 6B).

Experimentally, encoded slow initial translation does not increase gene expression; the opposite is true

Tuller et al. hypothesized that slow initial translation was adaptive, and improved efficiency of translation and gene expression by minimizing ribosome collisions. However, in the Spandrel hypothesis, slow initial translation is not generated by selection and is not adaptive. It might not have any effect on gene expression, but, if anything, slow translation might reduce gene expression. While informatics is wonderful, it is always nice to do an experiment, and in this section, we present direct experimental results regarding the effect of encoded slow, medium, and fast initial translation on gene expression.

We used a gene expression reporter based on EKD102414 (Methods and Materials). In this construct, GFP is the reporter, but for accuracy it is normalized against a divergently transcribed red fluorescent protein (Figure S6). Thus, GFP expression is reported as a GFP/RFP ratio. Although the reporter is GFP, the N-terminal region of this particular protein is derived from yeast HIS3, not GFP, and has little if any effect on the fluorescence of the GFP fused downstream. We used synonymous slow, medium, or fast codons to recode some of the codons in the first 41 amino acids of this GFP reporter to generate three reporters with slow, medium, or fast translation over the first 41 amino acids. The amino acid sequences of the three constructs were identical. The slow, medium, and fast average RRT values over the first 41 codons were 1.20, 1.04, and 0.93, respectively. That is, this SIT is slower than most natural SITs, and this FIT is faster than most natural FITs, but the difference is moderate.

As shown in Figure 7 (left three bars), the SIT did not improve expression of GFP, contrary to Tuller et al. In fact, the GFP with the SIT was expressed at only 71% of the level of the GFP with the FIT. It was surprising to us that the difference was this large—again, recoding was limited to codons within the first 41, and the protein sequences were identical.

Slow Initial Translation inhibits gene expression. Left 3 bars. A synthetic GFP was constructed with a leader amino acid sequence that had little effect on GFP. The leader sequence was recoded to give slow (SIT), medium (MIT) or fast (FIT) translation speed over the first 41 amino acids, without changing the amino acid sequence—i.e., the SIT, MIT and FIT had identical amino acid sequences, but different average RRTs. Each construct (SIT, MIT, FIT) was integrated in single copy at the ADE2 locus, and 25 independently-transformed strains were picked, and GFP fluorescence was measured for each, and the RFP-normalized mean was plotted. GFP was normalized to RFP expressed from the same reporter molecule, but RFP fluorescence hardly changed amongst the transformants, and non-normalized GFP would have given very similar results. Slower initial translation reduced gene expression. Right 3 bars. As above, but a Putative ribosome Collision Site (PCS) (CGA-CGG) was inserted between the leader and the GFP. Again, slower initial translation reduced gene expression.

Another possibility is that a SIT can protect against ribosome collisions when there is a site downstream that induces ribosome collisions. Sites thought to induce ribosome collisions include rare Arg-Arg codon pairs5, 15. We therefore introduced the codon pair CGA-CGG (replacing Asn-Asp, AAT-GAT) downstream of the first 41 amino acids, but still upstream of important GFP residues. Indeed, this single CGA-CGG codon pair, potentially inciting ribosome collisions, caused a large reduction--about 50%--in expression of GFP (Figure 7, right three bars). The reduction was about same in the SIT, MIT, and FIT constructs. In this case, with putative collision sites, the GFP with the SIT was expressed at only 67% of the level of the equivalent GFP with the FIT. That is, this SIT (a fairly extreme SIT) did not at all protect against the putative ribosome collisions—if anything, it made things worse. This result suggests there is no benefit to “queuing” ribosomes, if “queuing” even occurs. Instead, the fastest-translating gene once again gave the highest expression, despite the collision site.

Discussion

Tuller et al. found that the 5’ ends of genes are translated slowly because of the codons used at 5’ ends, and posited that this was a selective advantage because it somehow increased efficiency of translation. However, this theory predicts positive correlations between slow initial translation and high gene expression, and slow initial translation and high overall (that is, on the whole gene) ribosome density. In fact, by informatic analysis of existing data, we find the correlations are opposite to those predicted by the “Ramp” model. Most importantly, an experiment in which rates of initial translation are changed shows that faster initial translation causes higher gene expression, exactly opposite to the prediction of the “Ramp” hypothesis. We believe no ramp is needed.

Tuller et al. believed that a region of slow translation was encoded, using slowly-translated codons, and it is specifically this idea of encoded slow translation that we are addressing. However, ribosome profiling experiments done in the absence of cycloheximide still show an increased density of ribosome footprints near the 5’ end, independent of encoding3. This increased density could be due to slower translation at the 5’ end for some encoding-independent reason, or it could mean that some proportion of ribosomes fall off the mRNA as they progress3. Consistent with the latter idea, ribosome profiling studies show a general trend towards lower ribosome densities at more 3’ positions in translating mRNAs (Weinberg et al., their Figure S73). In any case, by direct experiment, we find that encoding a faster 5’ end improves gene expression.

We considered that the increased density of ribosomes at the 5’ end could be because some genes have additional ATG Start codons, sometimes upstream and sometimes downstream of the annotated Start, and translation of short open reading frames from these additional Start codons could contribute to ribosome density at the 5’ end. Using the program “Frameshift Detector”16 and ribosome profiling data, we quantitated the fraction of out-of-frame ribosomes both globally, and within the first 150 nucleotides of genes. We found the global proportion of out-of-frame ribosomes is about 13%, and the proportion of out-of-frame ribosomes in the first 150 nucleotides is about 14.6%. Although this increased out-of-frame ribosome presence of about 1.6% is highly significant (p ~ 10−28), it is not nearly big enough to explain the observed 5’ increase (~ 50%) in ribosome density (ref. 3, their Fig. 1C).

The observation that 5’ ends have low conservation, likely because of instability in evolution, provides a completely different explanation for the region of encoded slow initial translation. In this “Spandrel” hypothesis, N-termini frequently change in evolution, gathering new 5’ sequences de novo. These would contain all codons at similar frequencies—i.e., rare codons would not be especially rare. Although these rare codons would eventually be removed by selection, the fact that N-termini are relatively young means that this process might not be complete for all genes, and so some rare, slow codons still remain. These explain the initial region of encoded slow translation. This hypothesis is consistent with the observed correlations between slow initial translation and low gene expression; and slow initial translation and low ribosome density, and with the results of gene expression experiments. It is also consistent with the region of slow translation we observe at the 3’ ends of genes.

We have looked at the conservation of N-termini only in S. cerevisiae. However, Tuller et al. found that there is a region of encoded slow initial translation in genes of a wide variety of eukaryotes. We speculate that in these other cases, too, slow initial translation is a spandrel deriving from the turnover of 5’ ends. This in turn has implications for protein structure and evolution; for the interpretation of evolutionary sequence clocks; and for the rates of selection against rare codons.

It was surprising to us that recoding just the first 41 codons of the GFP fusion protein from slow to fast increased the level of GFP expression by so much—about 30%. These 40 codons were originally derived from the yeast HIS3 gene, and this increase in expression is roughly the proportional increase expected based on fully recoding the HIS3 gene to preferred codons17. Because this recoding from slow to fast tends to replace G/C rich codons with A/T rich codons, recoding from slow to fast may decrease the stability of RNA structures near the 5’ end, and increase the accessibility of the cap. Decreased stability of mRNA structures could be responsible for the increase in gene expression, consistent with studies in both yeast3, 18 and E. coli19.

We characterized both the 5’ and 3’ ends of genes. Qualitatively, results were similar: both 5’ and 3’ ends showed reduced conservation scores, increased proportions of rare codons, and slow translation, consistent with the Spandrel hypothesis. However, quantitatively, effects on rare codons and slow translation are much bigger at the 5’ end. We are puzzled by the asymmetry. Contributing to the asymmetry is the depletion of initiation codons from the 5’ end. These are fast codons; they are presumably depleted to prevent translation-initiation at the wrong place; and this effect is specific to the 5’ end. Also contributing are basic, arginine-rich sequences in N-terminal signal sequences; these Arg-rich sequences include many rare Arg codons, and this enrichment for N-terminal signal sequences is specific to the 5’ end. Still, even after taking these into account, quantitative effects seem bigger at the 5’ end. We speculate that although rare codons and slow translation are generated as spandrels, poor translation could be a selective advantage in genes that need to be expressed at low levels. Here, we are turning the hypothesis of Tuller et al. on its head: contrary to Tuller, slow initial translation seems to reduce gene expression, and yet this may be adaptive for some genes. Similarly, although architectural spandrels were originally generated from the need to support a lintel by an arch, nevertheless, there may eventually have been “selection” for them, as church-goers found them beautiful.

Finally, again, in “The Spandrels of San Marco …”, Gould and Lewontin 4 warned that not all biological phenomena are adaptive, and it is a mistake to assume that any particular characteristic of an organism must necessarily have been generated by natural selection. We believe the encoded slow initial translation of eukaryotic genes may be an example of this.

Methods and Materials

Statistical calculations of relative initial rate of translation

Bioinformatics were performed on protein-coding open reading frames (ORFs) of Saccharomyces cerevisiae downloaded from the Saccharomyces Genome Database (SGD) website, as last modified on April 22, 2021. Protein-coding ORFs annotated as dubious or pseudogenes were not included in analyses. All statistics were performed using The R Project for Statistical Computing. Translation speed was measured using the ribosome residence time (RRT) which is a metric of the occupancy of ribosomes on each sense codon within the A-site 6. The RRT values we used are shown in Table S1; these are modified from the original RRT results of Gardin et al. by inclusion of the ribosome profiling data of Jan et al.20.

Relative initial translation speed

Tuller et al.1 focused on a “ramp” of translation speed, where the first part of the gene has slow translation relative to the rest of the gene. The ramp thus refers to a rate. To quantitate this slow relative ramp for each gene, we calculated the average RRT for an initial window of the gene (e.g., 40 amino acids, see below), then divided by the average RRT of the rest of the gene. Thus, genes with a “slow ramp” have a ratio less than 1. We then took log2 of this ratio; genes with a slow ramp yield a negative number, and the more negative the number, the steeper the ramp.

For each gene, the relative initial translation speed (RIT) (explained above) was calculated across windows of the first 30, 40, 50 … and 100 codons, with all windows being statistically significant for slow translation. For these RIT calculations, the first (start) codon was omitted since all protein-coding genes in this dataset except Q0075 start with ATG, and ATG is one of the fastest codons, which would skew the RIT. Similarly, the last (stop) codon was omitted. Windows of the first 30, 40, 50 … and 100 codons each had statistically significant depletion in translation speed compared to the body of genes. We chose to focus on the first 40 codons. All genes shorter than 303 nucleotides (translated into 100 amino acids) were omitted from all analyses. For all RIT analyses, 328 out of 6022 ORFs were omitted leaving a dataset of 5694 ORFs.

Datasets

mRNA transcript readings, which we used as a proxy for gene expression, was acquired from Lipson et al. 200913. Genes with a read count less than 10 were omitted due to concerns about noise. Ribosome density measurements were acquired from Arava et al. 200312. These values were calculated as the number of ribosomes, detected on an mRNA, divided by the nucleotide length of the gene (including the stop codon).

Protein BLAST setup and diagnostics

S. cerevisiae proteins were downloaded from the SGD (last modified on April 22, 2021). Proteins derived from ORFs annotated as dubious or pseudogene were omitted from analyses. The Saccharomycotina (Taxonomy ID: 147537) protein sequences were downloaded from NCBI using the links:

https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Info&id=147537&lvl=3&lin=f&ke ep=1&srchmode=1&unlock

https://www.uniprot.org/uniprotkb?dir=ascend&query=(taxonomy_id:147537)&sort=organism_name

We downloaded and compiled the source databases from DDBJ, EMBL, Genbank, RefSeq, PIR, and UniProtKB. Duplicate sequences were deleted. To perform local BLAST, we downloaded the NCBI BLAST software (version 2.13.0+) and used RStudio as a wrapper to operate the software; all default BLAST parameters were selected, except that the number of alignments was changed to the maximum value of 1000000000. Local protein BLAST of every S. cerevisiae protein (5694 proteins, see above) was performed against all genomes of the subphylum Saccharomycotina (which includes the genus Saccharomyces), but omitting all species in the genus Saccharomyces. We eliminated submissions of duplicate species by limiting our database to the highest bitscores from sequence hits derived from each unique species. We were only interested in sequences that had high homology with queried S. cerevisiae proteins, so all hits with bitscores lower than 50 were omitted.

We wanted to compare conservation at the beginning of proteins with conservation at the middle and end of those same proteins. For this purpose, we split each S. cerevisiae protein into two halves (start to middle; middle to end), then blasted each half against all genomes in the subphylum Saccharomycotina (omitting Saccharomyces). We then calculated a “conservation score” (see below) for the first 40 amino acids of the protein, and, identically, for the first 40 amino acids of the second half of the protein. (We describe the first 40 amino acids of the second half of the protein as the “middle”, but in fact the region is displaced 20 amino acids C-terminal from the exact middle.) In a parallel way, a conservation score is calculated for the last 40 amino acids of each protein. For example, the length of Swi5 is 709 amino acids, and therefore BLASTs of the first half spanned from 1:354, and BLASTs for the second half spanned from 355:709. Conservation scores were calculated for residues 1:40 (beginning) (from the BLASTs of the first half of the protein), and 355:394 (middle) and 670:709 (end) (from BLASTs of the second half of the protein). In total, BLAST of the first half of all queried S. cerevisiae proteins yielded a total of 477,749 high homology (minimum of bitscore of 50) sequence matches across 817 unique Saccharomycotina species, whereas BLAST of the second half of each protein yielded a total of 486,871 high homology matches across 813 unique Saccharomycotina species.

With respect to the above procedure, we note that we are relying on the BLAST algorithm to find regions of homology. Homology would be somewhat more easily found in the middle of sequences than at the ends because of seeding issues. It is for this reason that we divided proteins in half, and used a BLAST with the second half of the protein to find homologies with the first 40 amino acids of the second half. That is, in this procedure, for the middle homologies, the algorithm is being asked to find homologies at the end of a sequence, exactly as is the case for the first 40 and last 40 amino acids.

Calculations of protein conservation scores and ratios

The general idea of the “Conservation Score” is that it represents the lengths of the regions of BLAST homology between the S. cerevisiae query and the Saccharomycotina subjects in the beginning, middle, and end 40-aminoacid windows. Each S. cerevisiae query sequence was separated into two equal halves, and then BLASTs were done on both halves against all of Saccharomycotina (omitting all submissions from the Saccharomyces genus). For all subject proteins with high homology over any region (i.e., a bitscore greater than 50), one finds the pair-wise regions of homology with a BLAST “Alignment Score” of 200 or more (red colored regions in the BLAST website “Graphic Summary”). The length of the high-homology region in the window of interest (but not the actual number of amino acid sequence matches within that region) contributes to the Conservation Score. That is, amino acids have to be within a region of homology found by BLAST in order to contribute to the score; even though all proteins begin with “M”, these only contribute to the conservation score if they are within a region of BLAST homology. The Conservation Score is the sum of (the length of the homology in the window, multiplied by the proportion of qualified subject proteins with that length of homology). Example Conservation Scores are shown in Table S2, and an example Conservation Score is calculated in Table S3. We only considered conservation scores from proteins that had homology with at least 40 unique species in Saccharomycotina. We also omitted proteins that were shorter than 100 amino acids. As shown in Table S2, conservation scores ranged from 0, meaning no BLAST homology region within the window of 40 amino acids for any qualifying homologue in Saccharomycotina, up to a maximum of 40, meaning that all qualifying homologues in Saccharomycotina had matches starting at the first amino acid. In total, protein conservation analyses used 3964 S. cerevisiae proteins with high homology hits for BLAST done on the first and second half of proteins (Figure 4).

Design of the fluorescent reporter gene constructs

We created a reporter gene based on reporter plasmid EKD102414. Briefly, a bidirectional galactose promoter simultaneously induces the expression of GFP and RFP in the presence of galactose, and we integrated the reporter into the yeast genome at the ADE2 locus. We recoded GFP to give the first 41 codons of GFP a slow initial translation speed (SIT), medium initial translation speed (MIT), or fast initial translation speed (FIT), while maintaining the same amino acid sequence (Figure S6, sequences in Table S4). We also designed three more constructs (Figure S6, sequences in Table S4) with a SIT, MIT, or FIT upstream of one of the slowest and rarest codon pairs, CGA-CGG (replacing AAT-GAT, Asn-Asp), which is known to greatly attenuate gene expression in living yeast21. Other rare codon pairs (CGA-CGA and CGA-CCG) have been shown to promote ribosome stalling15 so in Figure 6, right, CGA-CGG operates as a putative ribosome collision site (PCS). Instead of a PCS, the constructs in Figure 6, left, had AAT-GAT (Asn Asp), a frequent codon pair with above average translation speed. The Relative Initial Translation Speed scores of the constructs were: SIT was 0.208; MIT was 0.0004; FIT was −0.166; SIT+PCS was 0.198; MIT+PCS was −0.0109; and FIT+PCS was −0.177. (Note that the RIT scores of the constructs with the PCS change because the PCS makes the translation speed of the body of the gene slower; that is, the change is due to a change in the denominator.)

Yeast strains

The constructs were transformed into BY4741 (MATa; his3Δ1; leu2Δ0; met15Δ0; ura3Δ0). The reporter expresses MET15 allowing selection. Transformants were selected for Met+ on HULA plates (0.075g/L Histidine; 0.075g/L Uracil; 0.25g/L Leucine; 0.075g/L Adenine; 20g/L D-Glucose; 5g/L Ammonium Sulfate; 1.7g/L Yeast Nitrogen Base). The reporter gene integrates into the ADE2 locus, and thus successful transformants are ade2-delete which becomes red when grown on YPD plates; this was used as a secondary biological marker to confirm successful transformants. From each strain, a total of 25 Met+, Ade-, red transformants were chosen for each recoded GFP construct.

Flow Cytometry analysis

The strains were inoculated in liquid HULA media, with 2% galactose, until mid-log phase (around 10-14 hours) containing around 3 × 10^7 cells/mL. The strains were sonicated to separate cells and the strains were stored on ice (typically around an hour) until data collection. An LSR Fortessa Flow Cytometer was used to measure GFP levels (All Events FITC-A Mean) and RFP levels (All Events PE-Texas Red-A Mean) across 75,000 events. As a control, a strain lacking a fluorescent reporter was used. For all samples, GFP levels were normalized by RFP levels to reduce background fluorescence.

Statistical tests

None of our analyses made assumptions regarding the normality of the data. As such, we only performed nonparametric statistics. Wilcoxon signed-rank tests were done for relevant pairwise analyses; when necessary, p-values were corrected for multiple comparisons using the Holm–Bonferroni method. Spearman correlations were used. We used the Kolmogorov–Smirnov goodness of fit test to confirm that the three distributions were significantly different (p<0.001) for Figure 4.

Supporting information

eLife Assessment

This is an important contribution to the origins and translational consequences of the relatively low rate of translation elongation in the first ∼30-50 codons of genes in most organisms. The authors provide convincing evidence that the prevalence of rare codons in the first ~40 codons in yeast is due to the relatively recent evolution of these coding sequences, or of lower purifying selection operating on them, and that a preponderance of codons encoded by rare tRNAs near the N-terminus is not associated with higher translational efficiency in the manner proposed by the "translational ramp" hypothesis. There are, however, several incomplete aspects of the study: it neglects to discuss extensive published work providing different explanations for the slow rate of elongation at the start of open reading frames; it does not consider such effects in interpreting the reporter data produced for the current study; it does not resolve the apparent discrepancy posed by the lack of slow elongation observed despite poor sequence conservation of C-terminal coding regions; and it does not measure reporter mRNA levels to evaluate translational efficiencies.

Reviewer #1 (Public Review):

The manuscript by Sejour et al. is testing "translational ramp" model described previously by Tuller et al. in S. cerevisiae. Authors are using bioinformatics and reporter based experimental approaches to test whether "rare codons" in the first 40 codons of the gene coding sequences increase translation efficiency and regulate abundance of translation products in yeast cells. Authors conclude that "translation ramp" model does not have support using new set of reporters and bioinformatics analyses. The strength of bioinformatic evidence and experimental analyses of the rare codons insertion in the reporter make compelling case for authors claims. However major weakness of the manuscript is that authors do not take in account other confounding effects in their analyses as well as multiple previous studies that argue with "translation ramp" model. The existence of the early elongation ramp with "rare codons" was previously contested with local mRNA structure at the start codon, peptidyl-tRNA drop-off or interactions of the nascent peptide chain with exit channel of the ribosome models. All of these effects are not considered or discussed in the manuscript at this point. Such an authors approach makes the manuscript rather biased and short on discussing multiple other possible conclusions on reasons of slow translation elongation at the beginning of the protein synthesis.

Reviewer #2 (Public Review):

Tuller et al. first made the curious observation, that the first ∼30-50 codons in most organisms are encoded by scarce tRNAs and appear to be translated slower than the rest of the coding sequences (CDS). They speculated that this has evolved to pace ribosomes on CDS and prevent ribosome collisions during elongation - the "Ramp" hypothesis. Various aspects of this hypothesis, both factual and in terms of interpretating the results, have been challenged ever since. Sejour et al. present compelling results confirming the slower translation of the first ~40 codons in S. cerevisiae but providing alternative explanation for this phenomenon. Specifically, they show that the higher amino acid sequence divergence of N-terminal ends of proteins and accompanying lower purifying selection (perhaps the result of de novo evolution) is sufficient to explain the prevalence of rare slow codons in these regions. These results are an important contribution in understanding how aspects of evolution of protein coding regions can affect translation efficiency on these sequences and directly challenge the "Ramp" hypothesis proposed by Tuller et al.

I believe the data is presented clearly and the results generally justify the conclusions. I do have one specific concern related to interpretating the data. The authors show that the conservation score of the last 40 codons is not dissimilar to the conservation score of the first 40 (Fig. 4 A & C). They also show that the calculated translational speed of the first 40 codons is significantly lower than the rest of the CDS. At the same time, they show lack of statistically significant decrease of calculated translational speed for the last 40 codons (Figure S1). If the poor conservation of the first 40 codon explains the slower speed of their translation what is the authors' explanation for the absence of statistically significant reduction of calculated translational speed for the last 40 codons?

"Although the reporter is GFP, the N- terminal region of this particular protein is derived from yeast HIS3, not GFP, and has little if any effect on the fluorescence of the GFP fused downstream."

The statement above is logical and reasonable; however, it is not supported by any reference or control experiments. At the very least this fact should be explicitly acknowledged. Also, the RNA levels of reporters were not measured, which means it cannot be categorically concluded that the observed effect is due to changes of translational efficiency. This is an important caveat.

Author Response

The following is the authors’ response to the current reviews.

Reviewer #1 (Public Review):

The manuscript by Sejour et al. is testing "translational ramp" model described previously by Tuller et al. in S. cerevisiae. Authors are using bioinformatics and reporter based experimental approaches to test whether "rare codons" in the first 40 codons of the gene coding sequences increase translation efficiency and regulate abundance of translation products in yeast cells. Authors conclude that "translation ramp" model does not have support using a new set of reporters and bioinformatics analyses. The strength of bioinformatic evidence and experimental analyses (even very limited) of the rare codons insertion in the reporter make a compelling case for the authors claims. However the major weakness of the manuscript is that authors do not take into account other models that previously disputed "rare or slow codon" model of Tuller et al. and overstate their own results that are rather limited. This maintains to be the weak part of the manuscript even in the revised form.

We are glad the reviewer thinks our evidence makes “a compelling case for the authors claims”. This was our main aim, and we are satisfied with this.

The reviewer believes the major weakness of the manuscript is that we do not take into account other models and do not (see below) cite numerous other relevant papers. The reviewer made essentially the same criticism at the first review, at which time we looked quite hard for papers generally meeting the reviewer’s description. We found a few, which we incorporated here. Still, we did not find the body of evidence whose existence the reviewer implies. We are citing every study we know to be relevant, though of course we will have inadvertently missed some, given the huge body of literature. After the first round of review, we wrote “the reviewer did not give specific references, and, though we looked, we weren’t always sure which papers the reviewer had in mind.” We hoped the reviewer would provide citations. But only two citations are provided here, both to A. Kochetov, and these don’t seem central to the reviewer’s points.

The studies that authors do not mention argue with "translation ramp" model and show more thorough analyses of translation initiation to elongation transition as well as early elongation "slow down" in ribosome profiling data. Moreover several studies have used bioinformatical analyses to point out the evolution of N-terminal sequences in multiple model organisms including yeast, focusing on either upstream ORFs (uORFs) or already annotated ORFs. The authors did not mention multiple of these studies in their revised manuscript and did not comment on their own results in the context of these previous studies.

Mostly, we do not know to what papers the reviewer is referring. This may be our failing, but it would have helped if the reviewer had cited one of them. There are papers discussing the evolution of N-terminal sequences, but as far as we know, these do not discuss translation speed or codon usage. Of course, we may have missed some papers.

As such the authors approach to data presentation, writing and data discussion makes the manuscript rather biased, focused on criticizing Tuller et al. study and short on discussing multiple other possible reasons for slow translation elongation at the beginning of the protein synthesis. This all together makes the manuscript at the end very limited.

We think the reviewer may be considering our paper as being generally about translation speeds, whereas in our minds, it is not. This difference in views as to what the paper is “about” is perhaps causing friction. To us, it is indeed a limited paper. We are narrowly focused on the finding of Tuller that there is an enrichment of rare, slow codons at the 5’ end of genes, and we have sought an explanation of this particular fact. This is not a paper about rates of translation generally—it is a limited paper about the reason for the 5’ enrichment of rare, slow codons.

To expand on this, the encoded slow 5’ translation due to rare, slow codons (of Tuller et al.) is a small effect (1% to 3%). The possible unencoded slow 5’ translation of unknown mechanism discussed by some other papers (e.g., Weinberg et al. 2016, Shah et al. 2013) is a much larger effect (50% or more). Just from the different magnitudes, it seems likely these are different phenomena. And yet, despite the small size of the encoded effect, it is for some reason this paper by Tuller et al. that has captured the attention of the literature: as we point out below, Tuller et al. has been cited over 900 times. Partly because of the wide and continuing influence of this paper, it is worth specifically and narrowly addressing its findings.

Reviewer #2 (Public Review):

Tuller et al. first made the curious observation, that the first ∼30-50 codons in most organisms are encoded by scarce tRNAs and appear to be translated slower than the rest of the coding sequences (CDS). They speculated that this has evolved to pace ribosomes on CDS and prevent ribosome collisions during elongation - the "Ramp" hypothesis. Various aspects of this hypothesis, both factual and in terms of interpreting the results, have been challenged ever since. Sejour et al. present compelling results confirming the slower translation of the first ~40 codons in S. cerevisiae but providing an alternative explanation for this phenomenon. Specifically, they show that the higher amino acid sequence divergence of N-terminal ends of proteins and accompanying lower purifying selection (perhaps the result of de novo evolution) is sufficient to explain the prevalence of rare slow codons in these regions. These results are an important contribution in understanding how aspects of the evolution of protein coding regions can affect translation efficiency on these sequences and directly challenge the "Ramp" hypothesis proposed by Tuller et al.

I believe the data is presented clearly and the results generally justify the conclusions.

We thank the reviewer for his/her attention to the manuscript, and for his/her comments.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

As mentioned in the public review major weakness of the manuscript is the lack of analyses for confounding effects, overstatements of the results (using single amino acid sequence reporter) and the lack of discussion of previous work that argues against Tuller et al model. In my previous review I mentioned multiple other studies that addressed "slow codons" model in more detail.

No, the reviewer did not cite any specific studies.

While some of these studies are mentioned in the revised manuscript, authors are still rather biased and selective in their discussions. I should also point out that previous studies, that authors fail again to mention, were focused on either translation initiation, initiation to elongation transition or early elongation effects in relation to mRNA sequence, structure, codons as well as amino acid sequence. Also additional studies with bioinformatic analyses of N-terminal conservation and existence of start sites at the beginning of the protein sequences in multiple model organisms were also omitted.

Again, we do not know to what papers the reviewer is referring. But this sounds like a lot. Our paper is aimed at a specific, narrow topic: Why is there an excess of rare, slow codons in the 5’ region of genes? We are not trying to make general statements about all things affecting and affected by translation speed, we are just trying to explain the excess of rare, slow codons.

In general manuscript seems to be too much focused-on discussion of Tuller's paper . . .

Yes, we are focused on the Tuller findings, the excess of rare slow codons in 5’ regions.

. . . and arguing with the model that was already shown by multiple other studies to be limited and not correct.

We find it unsatisfactory that the reviewer states in a public review that there are multiple other studies showing that the Tuller model is not correct, and yet does not cite any of them. Furthermore, for the reviewer to say that Tuller et al. is “not correct” is too sweeping. The core finding of Tuller et al. was the excess of rare, slow codons in the 5’ regions of genes. We confirm this; we believe it is correct; we are not aware of any literature disputing this. Then, Tuller interpreted this as an adaptation to promote translational efficiency. On the interpretation, we disagree with Tuller. But if one is to disagree with this interpretation, one needs an alternative explanation of the fact of the excess rare, slow codons. Providing such an alternative explanation, and doing an experiment to distinguish the explanations, is our contribution. We are not aware of any other paper making our interpretation.

There are of course many papers that discuss various aspects of translation at the 5’ ends of genes, and we do cite quite a few such papers in our manuscript, though certainly not all. But papers of this general kind do not, and cannot, show that Tuller et al. is “not correct”. As far as we know, no paper provides an alternative explanation for the rare slow codons, and no paper does an experiment to modulate translation speed and look at the effect on gene expression. Notably, the slow translation phenomenon associated with the rare codons found by Tuller et al. is a very small effect—a change of about 1% to 3% of translation speed. Some other papers on translation speed are dealing with possible changes in the range of 50% or more. These are presumably some other phenomenon (if indeed they are even real changes in translation speed), and, whether they are true or not, the results and interpretations of Tuller et al. could still be true or not. Of course, if we knew of some previous paper showing the Tuller paper is not correct, we should and would cite it.

To expand on the current view of Tuller in the literature, Tuller et al. has been cited 956 times according to Google Scholar. This makes it an extremely influential paper. After finding Tuller et al. in Entrez Pubmed, one can look under “Cited by” and see the five most recent papers that cite Tuller et al. The five papers given on May 23 2024 were Bharti . . . Ignatova 2024; Uddin 2024; Khandia . . . Choudhary 2024; Love and Nair 2024; and Oelschlaeger 2024. We went through these five most recent papers that cite Tuller et al., and asked, did these authors cite the Tuller results as fully correct, or did they mention any doubts about the results? All five of the papers cited the Tuller results as fully correct, with no mention of any kind of doubt. For instance, Kandia et al. 2024 state “The slow “ramp” present at 5’ end of mRNA forms an optimal and robust means to reduce ribosomal traffic jams, thus minimizing the cost of protein expression40.”, while Oelschlaeger (2024) states “Slow translation ramps have also been described elsewhere and proposed to prevent traffic jams along the mRNA [51,52,53].” Although Uddin (2024) cited Tuller as fully correct, Uddin seemed to think (it is a little unclear) that Tuller found an enrichment of highly-used codons, opposite to the actual finding. The multiple contrary studies mentioned by the reviewer do not seem to have been very influential.

There are papers containing skepticism about the Tuller interpretation, and also papers with results that are difficult to reconcile in a common-sense way with the Tuller interpretation. But skepticism, and a difficulty to reconcile with common sense, are far from a demonstration that a paper is incorrect. Indeed, Tuller et al. may have been published in Cell, and may be so highly cited, exactly because the findings are counter-intuitive, colliding with common sense. Our contribution is to find a common-sense interpretation of the surprising but correct underlying fact of the 5’ enrichment of rare, slow codons.

Having wrote that in the previous review, I have to admit that Sejour et al manuscript in the main text has a minimal amount of novelty with experimental evidence, the conclusions are based on three reporters with and without stalling/collision sequence with the same amino acid sequence and varying codons. Some more novelty is seen in bioinformatic analyses of multiple yeast sequences and sequence conservation at the N-termini of proteins. However, even this part of the manuscript is not discussed fully and with correct comparison to previous studies. Authors, based on my previous comments discuss further experimental shortcomings in their new and "expanded" discussion but the use of a single reporter in this case cannot relate to all differences that may be coming from ORFs seen in complete yeast transcriptome. There are multiple studies that used more reporters with more than one amino-acid and mRNA sequence as well as with similar variation of the rare or common codons. The handwaving argument about the influence of all other mechanisms that can arise from different start sites, RNA structure, peptide interaction with exit channel, peptidyl-tRNA drop-off, eIF3 complex initiation-elongation association, and etc, is just pointing up to a manuscript that is more about bashing up Tuller's model and old paper than trying to make a concise story about their own results and discuss their study in plethora of studies that indicated multiple other models for slow early elongation.

We don’t understand why the reviewer is so grudging.

Discussion of the ribosome's collisions and potential impact of such scenario in the author's manuscript is left completely without citation, even though such work has relevant results to the author's conclusions and Tuller's model.

This is not true. We cite Dao Duc and Song (2018) “The impact of ribosomal interference, codon usage, and exit tunnel interactions on translation elongation rate variation.” PLoS Genet 14, and Tesina, . . . and Green (2020) “Molecular mechanism of translational stalling by inhibitory codon combinations and Poly(A) tracts. EMBO J., which are two excellent papers on this subject. We also cite Gamble et al. (2016), who found the underlying result, but at that time did not attribute it to ribosome collisions.

Previous studies (not cited) for example clearly indicate how the length from stalling sequence to start codon is related to ribosome collisions. Moreover such studies are pointing out differences in initiation vs elongation rates that may impact ribosome collisions and protein expression. Both of these topics would be very valuable in discussions of evolutionary changes in the current yeast ORFs. Not to mention that authors do not really discuss also possibilities for differences in 5'UTRs and uORFs in relation to downstream ORFs sequence and codon composition.

It is not clear to us that such papers are highly relevant to the issue on which we are working.

The argument about whether cycloheximide or not is doing 5' ribosome slowdown (lines 425-443) is just rambling about Weinberg's paper from 2016 without any real conclusion. In this section authors are just throwing down hypothesis that were more clearly explained in Weinberg's manuscript or shown experimentally in studies done after the Weinberg et al. paper was published.

Earlier, the reviewer had the criticism that “The studies that authors do not mention argue with "translation ramp" model and show more thorough analyses of translation initiation to elongation transition as well as early elongation "slow down" in ribosome profiling data.” The main study we know of dealing with these issues like these is that of Weinberg et al. 2016. In our opinion, this is a thoughtful paper on these issues. But now, at this point, the reviewer seems to criticize the fact that we do extensively cite results from Weinberg et al. It is true that there is no ultimate conclusion, but why there is no conclusion is a little bit interesting. Weinberg et al show that even in studies that do not use cycloheximide as the first step in ribosome profiling, there is some left-over high density of ribosomes near 5’ ends. But, all these ribosome profiling experiments do use cycloheximide at a later step in the procedure. Until someone does a ribosome profiling experiment without the use of any cycloheximide at any step, there will be no firm conclusion. This is not our fault—and also not the issue we are writing about. And, the reason this paragraph is in the manuscript at all is that the reviewer (we thought) had asked for something like this in the first review.

At the end, even in the limited novelty of evolutionary arguments about non-existing N-terminal conservation of codons or amino acids they fail to cite and discuss previous work by Kochetov (BioEssays, 2008 and NAR, 2011) which have additional explanation on evolution of N-terminal sequences in yeast, human or Drosophila.

These two papers of Dr. Kochetov’s have some relevance and we now cite them. These are the only papers cited by the reviewer in his/her two reviews.

Probably the reviewer would have preferred a paper on a different subject.

The following is the authors’ response to the original reviews.

Response to Reviewers:

We thank the reviewers for their comments, and their evident close reading of the manuscript. Generally, we agree with the reviewers on the strengths and weaknesses of our manuscript. Our revised manuscript has a more extensive discussion of alternative explanations for initial high ribosome density as seen by ribosome profiling, and which more specifically points out the limitations of our work.

As a preface to specific responses to the reviewers, we will say that we could divide observations of slow initial translation into two categories, which we will call “encoded slow codons”, and “increased ribosome density”. With respect to the first category, Tuller et al. documented initial “encoded slow codons”, that is, there is a statistical excess of rare, slowly-translated codons at the 5’ ends of genes. Although the size of this effect is small, statistical significance is extremely high, and the existence of this enrichment is not in any doubt. At first sight, this appears to be a strong indication of a preference for slow initial translation. In our opinion, our main contribution is to show that there is an alternative explanation for this initial enrichment of rare, slow codons—that they are a spandrel, a consequence of sequence plasticity at the 5’ (and 3’) ends of genes. The reviewers seem to generally agree with this, and we are not aware that any other work has provided an explanation for the 5’ enrichment of rare codons.

The second category of observations pertaining to slow initial translation is “increased ribosome density”. Early ribosome profiling studies used cycloheximide to arrest cell growth, and these studies showed a higher density of ribosomes near the 5’ end of genes than elsewhere. This high initial ribosome density helped motivate the paper of Tuller et al., though their finding of “encoded slow codons” could explain only a very small part of the increased ribosome density. More modern ribosome profiling studies do not use cycloheximide as the first step in arresting translation, and in these studies, the density of ribosomes near the 5’ end of genes is greatly reduced. And yet, there remains, even in the absence of cycloheximide at the first step, a significantly increased density of ribosomes near the 5’ end (e.g., Weinberg et al., 2016). (However, most or all of these studies do use cycloheximide at a later step in the protocol, and the possibility of a cycloheximide artefact is difficult to exclude.) Some of the reviewer’s concerns are that we do not explain the increased 5’ ribosome density seen by ribosome profiling. We agree; but we feel it is not the main point of our manuscript. In revision, we more extensively discuss other work on increased ribosome density, and more explicitly point out the limitations of our manuscript in this regard. We also note, though, that increased ribosome density is not a direct measure of translation speed—it can have other causes.

Specific Responses.

Reviewer 1 was concerned that we did not more fully discuss other work on possible reasons for slow initial translation. We discuss such work more extensively in our revision. However, as far as we know, none of this work proposes a reason for the 5’ enrichment of rare, slow codons, and this is the main point of our paper. Furthermore, it is not completely clear that there is any slow initial translation. The increase in ribosome density seen in flash-freeze ribosome profiling could be an artefact of the use of cycloheximide at the thaw step of the protocols; or it could be a real measure of high ribosome density that occurs for some other reason than slow translation (e.g., ribosomes might have low processivity at the 5’ end).

Reviewer 1 was also concerned about confounding effects in our reporter gene analysis of the effects of different codons on efficiency of translation. We have two comments. First, it is important to remember that although we changed codons in our reporters, we did not change any amino acids. We changed codons only to synonymous codons. Thus at least one of the reviewer’s possible confounding effects—interactions of the nascent peptide chain with the exit channel of the ribosome—does not apply. However, of course, the mRNA nucleotide sequence is altered, and this would cause a change in mRNA structure or abundance, which could matter. We agree this is a limitation to our approach. However, to fully address it, we feel it would be necessary to examine a really large number of quite different sequences, which is beyond the scope of this work. Furthermore, mRNAs with low secondary structure at the 5’ end probably have relatively high rates of initiation, and also relatively high rates of elongation, and it might be quite difficult to disentangle these. But in neither case is there an argument that slow initial translation is efficient. Accurate measurement of mRNA levels would be helpful, but would not disentangle rates of initiation from rates of elongation as causes of changes in expression.

Reviewer 2 was concerned that the conservation scores for the 5’ 40 amino acids, and the 3’ 40 amino acids were similar, but slow translation was only statistically significant for the 5’ 40 amino acids. As we say in the manuscript, we are also puzzled by this. We note that 3’ translation is statistically slow, if one looks over the last 100 amino acids. Our best effort at an explanation is a sort of reverse-Tuller explanation: that in the last 40 amino acids, the new slow codons created by genome plasticity are fairly quickly removed by purifying selection, but that in the first 40 amino acids, for genes that need to be expressed at low levels, purifying selection against slow codons is reduced, because poor translation is actually advantageous for these genes. To expand on this a bit, we feel that the 5000 or so proteins of the proteome have to be expressed in the correct stoichiometric ratios, and that poor translation can be a useful tool to help achieve this. In this explanation, slow translation at the 5’ end is bad for translation (in agreement with our reporter experiments), but can be good for the organism, when it occurs in front of a gene that needs to be expressed poorly. Whereas, in Tuller, slow translation at the 5’ end is good for translation.

Reviewer 2 wondered whether the N-terminal fusion peptide affects GFP fluorescence in our reporter. This specific reporter, with this N-terminus, has been characterized by Dean and Grayhack (2012), and by Gamble et al. (2016), and the idea that a super-folder GFP reporter is not greatly affected by N-terminal fusions is based on the work of Pedelacq (2006). None of these papers show whether this N-terminal fusion might have some effect, but together, they provide good reason to think that any effect would be small. These citations have been added.

Version 2

Abstract

Summary

Previously, Tuller et al. found that the first 30 to 50 codons of the genes of yeast and other eukaryotes are slightly enriched for rare codons, so are presumably translated somewhat slowly. They argued, based on informatics, that this initial slow translation “ramp” was adaptive; and increased efficiency of translation by queuing ribosomes to prevent collisions. Today, the translational speeds of different codons are known, and indeed rare codons are translated slowly. We re-examined the slow translation ramp. We confirm the finding that 5’ regions are enriched for rare codons. However, we also find that the 5’ ends of yeast genes are poorly conserved in evolution, suggesting that they are unstable and turn over relatively rapidly. When a new 5’ end forms de novo, it is likely to include codons that would otherwise be rare. Because evolution has had a relatively short time to select against these codons, 5’ ends are typically slightly enriched for rare, slow codons. Opposite to the expectation of Tuller et al., we show by direct experiment that genes with slowly translated codons at the 5’ end are expressed relatively poorly, and substituting faster codons improves expression. Further informatic studies suggest that for natural genes, slow 5’ ends are correlated with poor gene expression, opposite to the expectation of Tuller et al. Thus we conclude that slow 5’ translation is a “spandrel”; it is a non-adaptive consequence of something else, in this case the turnover of 5’ ends in evolution, and it does not improve translation.

Introduction

Tuller et al. 1 were interested in the idea that a slow translational ramp at the beginning of a gene might queue ribosomes in an orderly way, thereby preventing ribosome traffic jams and collisions. However, at that time, translation speeds for the 61 sense codons were not known from direct measurement. Therefore, as a proxy for codon translation speed, Tuller et al. devised a proxy speed measurement based on the tRNA-adaptation index (tAI), a measure of the abundance of each tRNA. The assumption is that codons recognized by more abundant tRNAs would be translated faster. Using this proxy, Tuller et al. found that in yeast and other eukaryotes, the first 30 to 100 codons are enriched for codons for which tRNAs are rare (generally, rare codons), and are presumably translated slowly. The size of the effect is small (about a 3% difference, Tuller et al. Figure 2C), but is statistically significant.

At the time of the work of Tuller et al., ribosome profiling had recently been developed2, and early ribosome profiling results showed a very high density of ribosomes near the 5’ end of the mRNA, consistent with very slow translation in this region, and the rare codons found by Tuller et al. could have contributed to this. However, later work showed that this 5’ high density of ribosomes was an artefact of the way cycloheximide was used to arrest translation in the original protocol3. With newer protocols for ribosome profiling, which use cycloheximide only at later steps, the region of 5’ high ribosome density largely, but not entirely, disappears3 (see Discussion).

Since then, many workers have used ribosome profiling to directly measure the speed of translation of individual codons (cited below). With such data in hand, we re-visited the issues addressed by Tuller et al. 1. On the one hand, our analyses confirm that the 5’ regions of genes are typically slightly enriched for rare codons, and these encodings likely slow translation. On the other hand, various aspects of the data led us to an alternative hypothesis, namely that the 5’ ends were turning over relatively rapidly in evolution; that these 5’ ends were therefore relatively young; and that selection had not yet succeeded in removing all the rare, slow codons initially present in the de novo 5’ ends. We did a direct experimental test of the effects of slow or fast initial translation. Exactly opposite to Tuller et al., we found that encoding slow initial translation resulted in lower protein production than fast initial translation. This continued to be true even when we placed ribosome collision sites inside the reporter gene. Thus a slow initial translation ramp, though present, does not improve gene expression.

It is natural to assume that the enrichment of slow codons near 5’ ends is a product of selection. However, as elegantly argued by Gould and Lewontin4 in their classic paper “The Spandrels of San Marco and the Panglossian Paradigm: A Critique of the Adaptionist Programme”, not all biological phenomena are adaptive, or even a direct product of selection. They argued from the example of a “spandrel”: in architecture, a triangular space created when an arch supports a lintel. There is no architectural role for spandrels as such; they are the indirect and inevitable result of the juxtaposition of two other functional architectural elements. We argue that the slightly slow initial translation of eukaryotic genes may likewise be a spandrel, a non-adaptive consequence of something else, the instability of 5’ ends in evolution.

Results

Calculations of encoded translation speed imply slow initial translation

Tuller et al. 1 used codon-specific tRNA abundance as a proxy to estimate speed of translation of codons. Since then, analysis of ribosome profiling data has yielded direct measurements of the translation speed of individual codons3, 5–11. Accordingly, we have repeated some of the work of Tuller et al., but using the Ribosome Residence Time (RRT, Methods and Materials, Table S1) 6 of each of the 61 sense codons as a measure of translation speed. We refer to “encoded translation speed” to specify that we are focusing purely on the effects of different codons on translation speed, and not on other factors that might differentially affect translation speed at different regions of the mRNA, such as secondary structure.

Using the RRT, encoded translation speeds were calculated in sliding windows across all coding ORFs from S. cerevisiae. (The start codon was omitted because it is constant across genes and is an unusually “fast” codon.) Consistent with Tuller et al., we find that the first 30-100 codons had lower calculated translation speeds than the rest of the gene (Figure 1). We focused our analyses on the first 40 codons for comparability to Tuller et al.; we call this the “Slow Initial Translation” region, or SIT. Although the tendency towards slow translation near the beginning of the gene is very highly statistically significant, the size of the effect is small. When we compare the first 40 codons of a gene to the rest of the same gene, we find that the average difference in encoded translation speed is about 1.2% (similar to 1), with a p-value < 0.001. For comparison, codons can vary in translation speed by about two-fold (Table S1), or perhaps as much as 6-fold (Weinberg et al., their Table S2, also shown in our Table S1).

Calculation of translation speed confirms slow initial translation (SIT). Translation speeds were calculated using Ribosome Residence Time (RRT) (Gardin et al. 2014) (Table S1 for RRT values) as a measure of codon-specific translation speed over 5694 S. cerevisiae ORFs. The horizontal line indicates average inverse RRT across all ORFs. The average speed in the first 40 amino acids is about 1.1% slower than in the rest of the gene (p < 0.001).

We observed a qualitatively similar, but smaller, encoded slow translation at the 3’ ends of genes, of borderline statistical significance (Figure S1).

Although on average genes are translated slowly near their 5’ ends, there is variability. Figure S2 shows the distribution of initial encoded translation speed for all yeast ORFs. About 57% of genes have a Slow Initial Translation (SIT) region, while the remainder have Fast Initial Translation (FIT).

Rare (slow) codons are enriched within the first 40 codons

Rare codons tend to be slow codons, and vice versa6 (Figure S3). But the correlation is not perfect—other things being equal, A/T-rich codons tend to be faster than G/C-rich codons, and codons with a wobble base at the third position tend to be faster than codons with the cognate base6. To better understand why gene beginnings are more slowly translated, we examined the relative usage of each of the 61 sense codons in the first 40 codons (Figure 2).

Codon usage in the Slow Initial Translation (SIT) region. A. Relative usage of each codon in the SIT versus the rest of the gene. The Y-axis shows the usage of each codon in the first 40 amino acids (omitting ATG) divided by its usage in the rest of gene. Each of the 61 sense codons is displayed along the X-axis, grouped by amino acid. Within each group, codons are ordered, left to right, from least frequent to most frequent over the whole transcriptome (i.e., the leftmost codon in each group is the rarest). Red arrows show the seven slowest codons by RRT, purple arrows show the seven rarest codons by total usage over the transcriptome, and blue shows Start and possible Start codons (ATG, ATA, ATC). Ratios above 1 indicate enrichment in the first 40 amino acids. Typically, the rarest codons for each amino acid are enriched. B. Absolute usage of each leucine codon in the SIT. The absolute usage frequency of each leucine codon is shown globally, and for the first 40 amino acids. Rare codons are still rare in the SIT, just not as rare as elsewhere. The same pattern holds for the other amino acids.

In the first 40 codons, we saw notable depletion of AUG (by nearly 50%), ATT, and ATC (Figure 2A). These are all fast codons, and their depletion would lead to slower translation. But AUG is the usual Start codon, and ATT and ATC can also be used as Start codons. Possibly these codons are depleted to reduce the possibility of translation initiating at the wrong place. We considered whether the SIT might be due to depletion of Start codons. We recalculated translation speeds after assigning AUG a neutral speed. On the one hand, this did reduce the difference in translation speed between the SIT and the rest of the gene by about 13% of the difference, but even so, the slow initial translation was highly significant. Thus, bias against Start codons contributes only modestly to slow translation.

We also saw enrichments of all six arginine codons (likely because of their use in N-terminal signal sequences, data not shown), especially the slow, rare codons CGA, CGC, and CGG. Slow, rare codons for proline (CCC, CCG), leucine (CTC), glycine (GGG) and cysteine (TGC) were also enriched (Figure 2A).

Rare codons are rare in the first 40 codons, just not as rare as elsewhere

Although the proportion of rare, slow codons in the SITs is relatively higher than in the body of genes, in absolute terms rare codons are still rare compared to more common synonymous codons (e.g., Figure 2B, Leu codons). That is, rare, slow codons are still strongly disfavored in the SIT, though they are less strongly disfavored than elsewhere. This was true for all rare codons.

Why are there relatively more rare, slow codons at 5’ ends?

Our analysis is consistent with that of Tuller et al. to the extent that we find a slight relative enrichment of rare, slow codons near the 5’ ends of coding regions. Tuller et al. interpret the slow translation ramp as an adaptation—they believe there is selection for slow codons near the 5’ end to enhance efficiency of translation. But there are other possibilities.

The Young Spandrel Hypothesis

We noticed (see below) that the N-termini of yeast genes are often poorly conserved, and otherwise highly homologous genes often vary at the N-termini between different closely-related species. This suggests a different idea: N-termini are unstable and variable in evolution. They form de novo from new DNA sequence, and so all codons may initially occur at similar frequencies. Since these N-termini are, on average, younger than the remainder of the gene, selection has worked on them for a shorter time. Therefore selection against rare, slow codons may be less complete, and N-termini, due to their relative youth, may still retain some extra rare codons.

In this idea, in contrast to Tuller et al., the slight excess of rare codons near N-termini is not adaptive; it is not at all a product of selection. Instead, in the words of Gould and Lewontin 4, it is a spandrel. It is a non-adaptive by-product of something else, in this case the evolutionary instability of N-termini. This idea is explored below.

Poor 5’ conservation is a feature of many yeast genes

To illustrate the idea, we picked example genes. We used protein-protein BLAST at NCBI to blast several query genes against species of the subphylum Saccharomycotina (but having subtracted out all of Saccharomyces). Figure 3 shows that for these chosen example genes, the middle portions of the proteins are highly conserved, but the N-termini are not. We suggest that these non-conserved N-terminal regions are young in evolution, and therefore likely contain an excess of rare codons.

The N-termini of proteins can vary in evolution. BLAST of four example S. cerevisiae proteins against proteins in the subphylum “Saccharomycotina” (taxid: 147537) (excluding Saccharomyces, taxid 4930) was performed. Top hits are shown. Red regions indicate homology with an alignment score > 200, while white indicates no detected homology (BLAST default parameters). Even though all hits have high to moderate homology towards the center of the protein, many have little or no homology at the N-terminus.

In the next section, we ask whether the findings of Figure 3 can be generalized to proteins of S. cerevisiae.

Method of scoring N-terminal conservation, and rationale for using Saccharomycotina

We continued our investigations into N-termini, protein conservation, and translation speed using a more systematic approach. We ran local protein-protein BLAST for all S. cerevisiae genes against a database consisting of the Saccharomycotina subphylum, omitting Saccharomyces cerevisiae. Saccharomycotina was chosen because almost every gene from S. cerevisiae has a recognizable, conserved homolog in almost every species in Saccharomycotina, and yet the evolutionary distances are long enough that there is considerable sequence variability. We excluded species of Saccharomyces as they are too closely related to S. cerevisiae, and they are also very numerous in the database, and would overwhelm results from the other members of Saccharomycotina. We believe, however, that this particular choice of subphylum does not greatly affect the final result.

Conservation at the N-terminus was calculated as the weighted proportion of yeast species with sequence matches (a match by default BLAST parameters) beginning in the first 40 amino acids. The lowest conservation score is 0 (no hits in the first 40 amino acids), whereas the highest conservation score is 40 indicating that every species had a match (default BLAST parameters) starting at the first amino acid (Table S2, S3). The length of genes is negatively correlated with the N-termini conservation score (rho= −.47; p<0.001)—that is, short genes tended to have more conserved N-termini.

N-termini are variable and poorly conserved

We measured protein conservation across more than 3,000 S. cerevisiae proteins with orthologues among 822 closely related yeasts from Saccharomycotina. For each protein, we developed a conservation score for the first 40 amino acids to represent the N-termini, an equivalent conservation score for the middle 40 amino acids, and an equivalent conservation score for the C-terminal 40 amino acids (Methods and Materials; Tables S2, S3).

Strikingly, the N-termini of S. cerevisiae orthologs had conservation scores that were much lower, and very differently distributed than the middle of the same orthologs (Figure 4). The first 40 amino acids had a flat distribution of protein conservation scores, indicating high levels of variability amongst these orthologs. That is, many of the orthologs had no detectable homology with the first 40 amino acids of the cerevisiae protein.

Conservation of S. cerevisiae proteins over the N-terminal, Middle, and C-terminal 40 amino acids. S. cerevisiae proteins were blasted against proteins of Saccharomycotina (excluding cerevisiae). “Conservation Scores” (Methods and Materials) were calculated for the N-terminal, Middle, and C-terminal 40 amino acids of the S. cerevisiae proteins. Scores range from 0 (no conservation) to 40 (perfect conservation). The frequency of each conservation score (3964 S. cerevisiae proteins) was plotted.

In contrast, the middle 40 amino acids were highly conserved, with conservations scores peaking at 40, the highest possible score. It Is evident that for the middle 40 amino acids, a large fraction of orthologs had a region of high homology to the S. cerevisiae protein, whereas this was not true for the N-termini. Finally, the last 40 amino acids had conservation scores similar to those of the first 40 amino acids, though a bit higher (more conserved) (see Figure S4 for a comparison). These results suggest both ends of the gene “breathe”, gaining and losing new sequences during evolution, whilst the middles of the same genes stay constant. Thus the ends of the genes are younger than the middles. At their first formation, they would likely have contained some rare codons, which selection may not yet have had time to remove.

5’ translation speeds positively correlate with 5’ conservation scores

If the “Young Spandrel” hypothesis is true, and slow 5’ translation is caused by evolutionary instability of 5’ ends, then there should be a correlation between encoded slow translation, and poor N-terminal conservation. Our model predicts the newest N-termini to have the lowest conservation and the rarest codons, and, vice versa, the termini with the rarest codons should have the lowest conservation scores. To test this, we ranked all genes by the conservation scores of their first 40 amino acids. We then divided this ranked list into thirds. For each of the thirds (i.e., the bottom, middle, and top conservation scores) we plotted the average relative initial translation speeds.

As shown in Figure 5A, the genes with the most poorly conserved N-termini also had the slowest initial translation, while the genes with the most conserved N-termini had the fasted initial translation, supporting the Spandrel hypothesis, and opposite to the Ramp hypothesis.

Slow Initial Translation is correlated with poor N-terminal conservation. A. Proteins were grouped by their N-terminal conservation scores (top, middle, and bottom thirds), and then the relative initial translation rate was plotted for each group. More conserved N-termini have faster initial translation. B. Proteins were grouped by their initial translation rate (top, middle, and bottom thirds), and then the N-terminal conservation scores were plotted for each group. Genes with faster initial translation have more conserved N-termini. Relative Initial Translation Speed is the log2 of (average RRT of the first 40 amino acids divided by the average RRT of the rest of the same gene) (Methods and Materials).

We also looked at the correlation in the other direction (Figure 5B). We ranked genes by their relative initial translation speed, and divided the ranked list into thirds, then plotted N-terminal conservation scores. Again the effects are correlated: genes with the slowest initial translation have the lowest N-terminal conservation scores. Thus, overall, there is a strong correlation between N-terminal instability (i.e., newness in evolution, low conservation scores) and slow initial translation (i.e., the presence of slow/rare codons).

The Ramp hypothesis is inconsistent with observations of ribosome density and gene expression

In the Tuller “Ramp” hypothesis, in which the purpose of the slow translational ramp is to queue ribosomes and prevent collisions, genes with the highest ribosome occupancy would be in the most danger of ribosome collisions, and would therefore presumably have pronounced SITs. SITs might not be necessary on genes with low ribosome density, since there would not be much danger of collision in any case. To test this, we used the Arava et al. 2003 dataset which measured the density of ribosomes on all S. cerevisiae mRNAs12 (Figure 6). We ranked genes by ribosome density, then grouped them in thirds. Opposite to the expectation from the Tuller et al. “Ramp” theory, the genes with the highest ribosome densities had the fastest initial translation, whereas the genes with the lowest ribosome densities had the slowest initial translation (Figure 6C). While these findings are opposite to the expectation of the “Ramp” theory, they are consistent with the spandrel theory, because genes with high ribosome density would be subject to more intense selection against slow codons, thus leading to faster 5’ ends. Furthermore, analysis of Conservation Scores on the same genes showed that the genes with the lowest ribosome densities also had the lowest Conservation Scores (Figure 6D), as predicted by the Spandrel hypothesis.

Genes with high levels of expression, and high ribosome densities, generally have rapidly-translated N-termini, and high N-terminal conservation scores. A and B. Genes were grouped by expression level (bottom, middle, and top)(except that genes with fewer than 10 read-counts were omitted to reduce noise) (Lipson et al. 2009). In A, the initial translation rate is shown; in B, the conservation scores are shown. The correlation between speed and transcript abundance fails for the bottom third of genes; possibly these are genes expressed at high levels under other conditions (e.g., meiosis and sporulation). C and D. Genes were grouped by ribosome density (Arava et al. 2003) as a measure of intensity of translation. In C, the initial translation rate is shown; in D, the conservation scores are shown. High ribosome density correlates with high initial translation speed and high conservation score.

Similarly, in the “Ramp” hypothesis, ribosome collisions on highly-expressed genes would presumably have more serious consequences for the cell than collisions on poorly-expressed genes, and so highly expressed genes ought to have the most pronounced SITs. To test this, we used the Lipson et al. 2009 transcriptome dataset which measured the number of mRNA transcripts for S. cerevisiae genes13. Again, we ranked genes by expression, then grouped them by thirds. Exactly contrary to the “Ramp” hypothesis, we found that genes with the highest expression had the fastest initial translation (Figure 6A). In this analysis, the middle and bottom genes are not significantly different from each other; possibly some of the poorly expressed genes are inducible genes that would be highly expressed under some other condition (e.g., the sporulation genes, the GAL genes). In addition, the most highly expressed genes had the highest conservation scores, consistent with the Spandrel hypothesis (Figure 6B).

Experimentally, encoded slow initial translation does not increase gene expression; the opposite is true

Tuller et al. hypothesized that slow initial translation was adaptive, and improved efficiency of translation and gene expression by minimizing ribosome collisions. However, in the Spandrel hypothesis, slow initial translation is not generated by selection and is not adaptive. It might not have any effect on gene expression, but, if anything, slow translation might reduce gene expression. While informatics is wonderful, it is always nice to do an experiment, and in this section, we present direct experimental results regarding the effect of encoded slow, medium, and fast initial translation on gene expression.

We used a gene expression reporter based on EKD102414 (Methods and Materials). In this construct, GFP is the reporter, but for accuracy it is normalized against a divergently transcribed red fluorescent protein (Figure S6). Thus, GFP expression is reported as a GFP/RFP ratio. Although the reporter is GFP, the N-terminal region of this particular protein is derived from yeast HIS3, not GFP, and likely has little if any effect on the fluorescence of the GFP fused downstream15–17. We used synonymous slow, medium, or fast codons to recode some of the codons in the first 41 amino acids of this GFP reporter to generate three reporters with slow, medium, or fast translation over the first 41 amino acids. We emphasize that the amino acid sequences of the three constructs were identical. The slow, medium, and fast average RRT values over the first 41 codons were 1.20, 1.04, and 0.93, respectively. That is, this SIT is slower than most natural SITs, and this FIT is faster than most natural FITs, but the difference is moderate.

As shown in Figure 7 (left three bars), the SIT did not improve expression of GFP, contrary to Tuller et al. In fact, the GFP with the SIT was expressed at only 71% of the level of the GFP with the FIT. It was surprising to us that the difference was this large—again, recoding was limited to codons within the first 41, and the protein sequences were identical.

Slow Initial Translation inhibits gene expression. Left 3 bars. A synthetic GFP was constructed with a leader amino acid sequence that had little effect on GFP. The leader sequence was recoded to give slow (SIT), medium (MIT) or fast (FIT) translation speed over the first 41 amino acids, without chang-ing the amino acid sequence—i.e., the SIT, MIT and FIT had identical amino acid sequences, but different average RRTs. Each construct (SIT, MIT, FIT) was integrated in single copy at the ADE2 locus, and 25 independently-transformed strains were picked, and GFP fluorescence was measured for each, and the RFP-normalized mean was plotted. Numerical values were: SIT, 1.66; MIT, 1.80, FIT, 2.29. GFP was normalized to RFP expressed from the same reporter molecule, but RFP fluorescence hardly changed amongst the transformants, and non-normalized GFP would have given very similar results. Slower initial translation reduced gene expression. Right 3 bars. As above, but a Putative ribosome Collision Site (PCS) (CGA-CGG) was inserted between the leader and the GFP. Again, slower initial translation reduced gene expression. Values were: SIT:PCS, 0.69, MIT:PCS, 0.74, FIT:PCS, 0.99.

Another possibility is that a SIT can protect against ribosome collisions when there is a site downstream that induces ribosome collisions. Sites thought to induce ribosome collisions include rare Arg-Arg codon pairs5,18. We therefore introduced the codon pair CGA-CGG (replacing Asn-Asp, AAT-GAT) downstream of the first 41 amino acids, but still upstream of important GFP residues. Indeed, this single CGA-CGG codon pair, potentially inciting ribosome collisions, caused a large reduction--about 50%--in expression of GFP (Figure 7, right three bars). The reduction was about same in the SIT, MIT, and FIT constructs. In this case, with putative collision sites, the GFP with the SIT was expressed at only 67% of the level of the equivalent GFP with the FIT. That is, this SIT (a fairly extreme SIT) did not at all protect against the putative ribosome collisions—if anything, it made things worse. This result suggests there is no benefit to “queuing” ribosomes, if “queuing” even occurs. Instead, the fastest-translating gene once again gave the highest expression, and the highest relative expression, despite the collision site.

Discussion

Tuller et al. found that the 5’ ends of genes are translated slowly because of the codons used at 5’ ends, and posited that this was a selective advantage because it somehow increased efficiency of translation. However, this theory predicts positive correlations between slow initial translation and high gene expression, and slow initial translation and high overall (that is, on the whole gene) ribosome density. In fact, by informatic analysis of existing data, we find the correlations are opposite to those predicted by the “Ramp” model. Most importantly, an experiment in which rates of initial translation were changed shows that faster initial translation causes higher gene expression, exactly opposite to the prediction of the “Ramp” hypothesis. We believe no ramp is needed.

Tuller et al. believed that a region of slow translation is encoded, using slowly-translated codons, and it is specifically this idea of encoded slow translation that we are addressing. However, in addition to “encoded” slow translation, there is some evidence for slow translation at 5’ ends by other, unknown mechanisms. Ribosome profiling experiments show an increased density of ribosome footprints near the 5’ end, independent of encoding3, and a high density of footprints could be due to slow translation. It is now known that much of the reason for increased density of footprints in early studies was the use of cycloheximide as the first step in the original ribosome footprinting protocols. Addition of cycloheximide to growing cultures allowed ribosomes to initiate at Start codons, but did not allow elongation, hence there was a pile-up of ribosomes near the 5’ end. More recently, flash-freezing, rather than cycloheximide, has been used as the first step in stopping translation, and even in these studies, there is about a 50% increase in ribosome density near 5’ ends3. And yet, even in these flash-freezing protocols, cycloheximide is used in a later step, when extracts are thawed, to prevent elongation, and this cycloheximide usage at a second step could again result in an artefactual increase in ribosome density at 5’ ends. Alternatively, the increased density of ribosomes at 5’ ends could mean that some proportion of ribosomes fall off the mRNA as they progress3. Consistent with the latter idea, ribosome profiling studies show a general trend towards lower ribosome densities at more 3’ positions in translating mRNAs (Weinberg et al., their Figure S73), and studies using other experimental approaches have shown a general decrease in ribosome number or density as one progresses along a gene 19,20. For these reasons, it is not clear that there is any slow translation in the 5’ regions of genes.

We considered that an increased density of ribosomes at the 5’ end could be because some genes have additional ATG Start codons, sometimes upstream and sometimes downstream of the annotated Start, and translation of short open reading frames from these additional Start codons could contribute to ribosome density at the 5’ end. Using the program “Frameshift Detector”21 and ribosome profiling data, we quantitated the fraction of out-of-frame ribosomes both globally, and within the first 150 nucleotides of genes. We found the global proportion of out-of-frame ribosomes is about 13%, and the proportion of out-of-frame ribosomes in the first 150 nucleotides is about 14.5%. Although this increased out-of-frame ribosome presence of about 1.5% is highly significant (p ∼ 10-28), it is not nearly big enough to explain the observed 5’ increase (∼ 50%) in ribosome density (ref. 3, their Fig. 1C).

In any case, by direct experiment, we find that encoding a faster 5’ end using fast synonymous codons improves gene expression. In particular, even when a ribosome collision site was placed downstream, the effect of the collision site was not at all ameliorated by encoded slow translation upstream of the collision site. This seems strong evidence against the idea that slow initial translation is a defense against collisions. The basis of the idea that slow initial translation could possibly be a defense against collisions is not clear to us. Regions of slow translation would not affect the gaps between ribosomes, if measured as times, and especially not if measured at a constant finish-line, such as a putative collision site.

A potential weakness in our GFP reporter experiment is that the mRNA sequences are necessarily different, and so there are different mRNA structures. We achieved fast and slow sequences by recoding with synonymous codons, so amino acids are identical, and there are no issues of, e.g., co-translational protein folding, or interaction with the ribosome exit tunnel. But of course the RNA structures will be at least slightly different, and RNA structures at the 5’ end are known to affect translation initiation3,22–27. Generally, more open mRNA structures are more favorable for translation initiation, and also for translation speed. To fully resolve this would likely require the recoded GFP reporter experiment to be redone many times, starting from different underlying sequences, and even then, it might be hard to disentangle effects of open mRNA structure on initiation from effects of fast translation. But whether the increased gene expression we see for the “fast” encoding is being generated mainly by fast translation, or by efficient initiation, in neither case is there an argument that slow translation is efficient, or that it protects against collisions.

The observation that 5’ ends have low conservation, likely because of instability in evolution, provides a completely different explanation for the region of encoded slow initial translation. In this “Spandrel” hypothesis, N-termini frequently change in evolution, gathering new 5’ sequences de novo. These would contain all codons at similar frequencies—i.e., rare codons would not be especially rare. Although these rare codons would eventually be removed by selection, the fact that N-termini are relatively young means that this process might not be complete for all genes, and so some rare, slow codons still remain. These explain the initial region of encoded slow translation. This hypothesis is highly consistent with the observed correlations between slow initial translation and low gene expression; and slow initial translation and low ribosome density, and with the results of gene expression experiments. It is also consistent with the region of slightly slow translation we observe at the 3’ ends of genes.

We have looked at the conservation of N-termini only in S. cerevisiae. However, Tuller et al. found that there is a region of encoded slow initial translation in genes of a wide variety of eukaryotes. We speculate that in these other cases, too, slow initial translation is a spandrel deriving from the turnover of 5’ ends. This in turn has implications for protein structure and evolution; for the interpretation of evolutionary sequence clocks; and for the rates of selection against rare codons. We note that Bricout et al.28 have also recently found that N- and C-termini of proteins evolve faster than the middles, although their perspective is different from ours.

It was surprising to us that recoding just the first 41 codons of the GFP fusion protein from slow to fast increased the level of GFP expression by so much—about 30%. These 41 codons were originally derived from the yeast HIS3 gene, and this increase in expression is roughly the proportional increase expected based on fully recoding the HIS3 gene to preferred codons29. Because this recoding from slow to fast tends to replace G/C rich codons with A/T rich codons, recoding from slow to fast may decrease the stability of RNA structures near the 5’ end, and increase the accessibility of the cap. Decreased stability of mRNA structures could be responsible for the increase in gene expression, consistent with studies in both yeast3,23 and E. coli26.

We characterized both the 5’ and 3’ ends of genes. Qualitatively, results were similar: both 5’ and 3’ ends showed reduced conservation scores, increased proportions of rare codons, and slow translation, consistent with the Spandrel hypothesis. However, quantitatively, effects on rare codons and slow translation are much bigger at the 5’ end. We are puzzled by the asymmetry. Contributing to the asymmetry is the depletion of initiation codons from the 5’ end. These are fast codons; they are presumably depleted to prevent translation-initiation at the wrong place; and this effect is specific to the 5’ end. Also contributing are basic, arginine-rich sequences in N-terminal signal sequences; these Arg-rich sequences include many rare Arg codons, and this enrichment for N-terminal signal sequences is specific to the 5’ end. Still, even after taking these into account, quantitative effects seem bigger at the 5’ end. We speculate that although rare codons and slow translation are generated and maintained for a time as spandrels, poor translation could be a selective advantage in genes that need to be expressed at low levels. Here, we are turning the hypothesis of Tuller et al. on its head: contrary to Tuller, slow initial translation seems to reduce gene expression, and yet this may be adaptive for some genes. Similarly, although architectural spandrels were originally generated from the need to support a lintel by an arch, nevertheless, there may eventually have been “selection” for them, as church-goers found them beautiful.

We believe our study has three main limitations or uncertainties. First, in our GFP reporter experiments suggesting that a slowly-translated N-terminal region reduces gene expression, we did not address the possibility that changes in mRNA structure could have been responsible for this. We believe that experiments with a multitude of reporter sequences would be required to address this issue. Second, we found somewhat different results at the 5’ and 3’ ends of genes: both ends show poor evolutionary conservation, but the enrichment of rare codons is more pronounced at the 5’ end. Although we suggest this could be due to selection for poor translation at the 5’ end (see above), this is only a suggestion, and we feel the reasons for the difference between the 5’ and 3’ ends are unresolved. Third, ribosome profiling studies show high ribosome densities near the 5’ end of the mRNA which are not nearly fully accounted for by rare codons3 (and see Discussion above); although we discuss (above) several possibilities for these high 5’ ribosome densities, ultimately this issue also remains unresolved.

Finally, again, in “The Spandrels of San Marco …”, Gould and Lewontin4 warned that not all biological phenomena are adaptive, and it is a mistake to assume that any particular characteristic of an organism must necessarily have been generated by natural selection. We believe the encoded slow initial translation of eukaryotic genes may be an example of this.

Methods and Materials

Statistical calculations of relative initial rate of translation

Bioinformatics were performed on protein-coding open reading frames (ORFs) of Saccharomyces cerevisiae downloaded from the Saccharomyces Genome Database (SGD) website, as last modified on April 22, 2021. Protein-coding ORFs annotated as dubious or pseudogenes were not included in analyses. All statistics were performed using The R Project for Statistical Computing. Translation speed was measured using the ribosome residence time (RRT) which is a metric of the occupancy of ribosomes on each sense codon within the A-site 6. The RRT values we used are shown in Table S1; these are modified from the original RRT results of Gardin et al. by inclusion of the ribosome profiling data of Jan et al.30.

Relative initial translation speed

Tuller et al.1 focused on a “ramp” of translation speed, where the first part of the gene has slow translation relative to the rest of the gene. The ramp thus refers to a rate. To quantitate this slow relative ramp for each gene, we calculated the average RRT for an initial window of the gene (e.g., 40 amino acids, see below), then divided by the average RRT of the rest of the gene. Thus, genes with a “slow ramp” have a ratio less than 1. We then took log2 of this ratio; genes with a slow ramp yield a negative number, and the more negative the number, the steeper the ramp.

For each gene, the relative initial translation speed (RIT) (explained above) was calculated across windows of the first 30, 40, 50 … and 100 codons, with all windows being statistically significant for slow translation. For these RIT calculations, the first (start) codon was omitted since all protein-coding genes in this dataset except Q0075 start with ATG, and ATG is one of the fastest codons, which would skew the RIT. Similarly, the last (stop) codon was omitted. Windows of the first 30, 40, 50 … and 100 codons each had statistically significant depletion in translation speed compared to the body of genes. We chose to focus on the first 40 codons. All genes shorter than 303 nucleotides (translated into 100 amino acids) were omitted from all analyses. For all RIT analyses, 328 out of 6022 ORFs were omitted leaving a dataset of 5694 ORFs.

Datasets

mRNA transcript readings, which we used as a proxy for gene expression, was acquired from Lipson et al. 200913. Genes with a read count less than 10 were omitted due to concerns about noise. Ribosome density measurements were acquired from Arava et al. 200312. These values were calculated as the number of ribosomes, detected on an mRNA, divided by the nucleotide length of the gene (including the stop codon).

Protein BLAST setup and diagnostics

S. cerevisiae proteins were downloaded from the SGD (last modified on April 22, 2021). Proteins derived from ORFs annotated as dubious or pseudogene were omitted from analyses. The Saccharomycotina (Taxonomy ID: 147537) protein sequences were downloaded from NCBI using the links:

https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Info&id=147537&lvl=3&lin=f&keep=1&srchmode=1&unlock

https://www.uniprot.org/uniprotkb?dir=ascend&query=(taxonomy_id:147537)&sort=organism_name

We downloaded and compiled the source databases from DDBJ, EMBL, Genbank, RefSeq, PIR, and UniProtKB. Duplicate sequences were deleted. To perform local BLAST, we downloaded the NCBI BLAST software (version 2.13.0+) and used RStudio as a wrapper to operate the software; all default BLAST parameters were selected, except that the number of alignments was changed to the maximum value of 1000000000. Local protein BLAST of every S. cerevisiae protein (5694 proteins, see above) was performed against all genomes of the subphylum Saccharomycotina, but omitting all species in the genus Saccharomyces (net, 822 genomes). We eliminated submissions of duplicate species by limiting our database to the highest bitscores from sequence hits derived from each unique species. We were only interested in sequences that had high homology with queried S. cerevisiae proteins, so all hits with bitscores lower than 50 were omitted.

We wanted to compare conservation at the beginning of proteins with conservation at the middle and end of those same proteins. For this purpose, we split each S. cerevisiae protein into two halves (start to middle; middle to end), then blasted each half against all genomes in the subphylum Saccharomycotina (omitting Saccharomyces). We then calculated a “conservation score” (see below) for the first 40 amino acids of the protein, and, identically, for the first 40 amino acids of the second half of the protein. (We describe the first 40 amino acids of the second half of the protein as the “middle”, but in fact the region is displaced 20 amino acids C-terminal from the exact middle.) In a parallel way, a conservation score is calculated for the last 40 amino acids of each protein. For example, the length of Swi5 is 709 amino acids, and therefore BLASTs of the first half spanned from 1:354, and BLASTs for the second half spanned from 355:709. Conservation scores were calculated for residues 1:40 (beginning) (from the BLASTs of the first half of the protein), and 355:394 (middle) and 670:709 (end) (from BLASTs of the second half of the protein). In total, BLAST of the first half of all queried S. cerevisiae proteins yielded a total of 477,749 high homology (minimum of bitscore of 50) sequence matches across 816 unique Saccharomycotina species, whereas BLAST of the second half of each protein yielded a total of 487,022 high homology matches across 812 unique Saccharomycotina species.

With respect to the above procedure, we note that we are relying on the BLAST algorithm to find regions of homology. Homology would be somewhat more easily found in the middle of sequences than at the ends because of seeding issues. It is for this reason that we divided proteins in half, and used a BLAST with the second half of the protein to find homologies with the first 40 amino acids of the second half. That is, in this procedure, for the middle homologies, the algorithm is being asked to find homologies at the end of a sequence, exactly as is the case for the first 40 and last 40 amino acids.

Calculations of protein conservation scores and ratios

The general idea of the “Conservation Score” is that it represents the lengths of the regions of BLAST homology between the S. cerevisiae query and the Saccharomycotina subjects in the beginning, middle, and end 40-aminoacid windows. Each S. cerevisiae query sequence was separated into two equal halves, and then BLASTs were done on both halves against all of Saccharomycotina (omitting all submissions from the Saccharomyces genus). For all subject proteins with high homology over any region (i.e., a bitscore greater than 50), one finds the pair-wise regions of homology with a BLAST “Alignment Score” of 200 or more (red colored regions in the BLAST website “Graphic Summary”). The length of the high-homology region in the window of interest (but not the actual number of amino acid sequence matches within that region) contributes to the Conservation Score. That is, amino acids have to be within a region of homology found by BLAST in order to contribute to the score; even though all proteins begin with “M”, these only contribute to the conservation score if they are within a region of BLAST homology. The Conservation Score is the sum of (the length of the homology in the window, multiplied by the proportion of qualified subject proteins with that length of homology). Example Conservation Scores are shown in Table S2, and an example Conservation Score is calculated in Table S3. We only considered conservation scores from proteins that had homology with at least 40 unique species in Saccharomycotina. We also omitted proteins that were shorter than 100 amino acids. As shown in Table S2, conservation scores ranged from 0, meaning no BLAST homology region within the window of 40 amino acids for any qualifying homologue in Saccharomycotina, up to a maximum of 40, meaning that all qualifying homologues in Saccharomycotina had matches starting at the first amino acid. In total, protein conservation analyses used 3964 S. cerevisiae proteins with high homology hits for BLAST done on the first and second half of proteins (Figure 4).

Code

We provide the custom R code written for this project as a text file, Appendix 1.

Design of the fluorescent reporter gene constructs

We created a reporter gene based on reporter plasmid EKD102414. Briefly, a bidirectional galactose promoter simultaneously induces the expression of GFP and RFP in the presence of galactose, and we integrated the reporter into the yeast genome at the ADE2 locus. We recoded GFP to give the first 41 codons of GFP a slow initial translation speed (SIT), medium initial translation speed (MIT), or fast initial translation speed (FIT), while maintaining the same amino acid sequence (Figure S6, sequences in Table S4). We also designed three more constructs (Figure S6, sequences in Table S4) with a SIT, MIT, or FIT upstream of one of the slowest and rarest codon pairs, CGA-CGG (replacing AAT-GAT, Asn-Asp), which is known to greatly attenuate gene expression in living yeast16. Other rare codon pairs (CGA-CGA and CGA-CCG) have been shown to promote ribosome stalling18 so in Figure 6, right, CGA-CGG operates as a putative ribosome collision site (PCS). Instead of a PCS, the constructs in Figure 6, left, had AAT-GAT (Asn Asp), a frequent codon pair with above average translation speed. The Relative Initial Translation Speed scores of the constructs were: SIT was 0.208; MIT was −.0004; FIT was −.166; SIT+PCS was 0.198; MIT+PCS was −.0109; and FIT+PCS was −.177. (Note that the RIT scores of the constructs with the PCS change because the PCS makes the translation speed of the body of the gene slower; that is, the change is due to a change in the denominator.)

Yeast strains

The constructs were transformed into BY4741 (MATa; his3Δ1; leu2Δ0; met15Δ0; ura3Δ0). The reporter expresses MET15 allowing selection. Transformants were selected for Met+ on HULA plates (0.075g/L Histidine; 0.075g/L Uracil; 0.25g/L Leucine; 0.075g/L Adenine; 20g/L D-Glucose; 5g/L Ammonium Sulfate; 1.7g/L Yeast Nitrogen Base). The reporter gene integrates into the ADE2 locus, and thus successful transformants are ade2-delete which becomes red when grown on YPD plates; this was used as a secondary biological marker to confirm successful transformants. From each strain, a total of 25 Met+, Ade-, red transformants were chosen for each recoded GFP construct.

Flow Cytometry analysis

The strains were inoculated in liquid HULA media, with 2% galactose, until mid-log phase (around 10-14 hours) containing around 3 x 10^7 cells/mL. The strains were sonicated to separate cells and the strains were stored on ice (typically around an hour) until data collection. An LSR Fortessa Flow Cytometer was used to measure GFP levels (All Events FITC-A Mean) and RFP levels (All Events PE-Texas Red-A Mean) across 75,000 events. As a control, a strain lacking a fluorescent reporter was used. For all samples, GFP levels were normalized by RFP levels to reduce background fluorescence.

Statistical tests

None of our analyses made assumptions regarding the normality of the data. As such, we only performed nonparametric statistics. Wilcoxon signed-rank tests were done for relevant pairwise analyses; when necessary, p-values were corrected for multiple comparisons using the Holm–Bonferroni method. Spearman correlations were used. We used the Kolmogorov–Smirnov goodness of fit test to confirm that the three distributions were significantly different (p<0.001) for Figure 4.

Supporting information